Vertical and oblique transmission with fluctuating transmission

### 2018-05-29 ms/ms_sunnyvale.tex
real roots for some cases, need to finish

\documentclass[14pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

@@

Important modes of cultural transmission are vertical (from parent to offspring), horizontal (between peers), and oblique (from non-parental adults to offspring) transmission. 
\citet[ch.~3]{Cavalli-Sforza1981} first introduced models in which a specific trait is transmitted either vertically or obliquely.
Mathematical models of cultural evolution may include one or more modes of cultural transmission and in most of these models the mode of transmission by individuals is fixed; that is, it does not vary over time.
For example, \cite{Feldman1996}, \cite{Wakano2004}, \cite{Aoki2005}, and \citet{McElreath2008} focused on competition between individual learning, innate learning, and social learning; \citet{Fogarty2017} compared the effects of different oblique mechanisms (random, success-biased, best-of-k, one-to-many) on the cultural richness and diversity of the population; and \citet{Aoki2012} modeled scenarios in which individual and social learning occur during separate stages in life.
In these studies, the cultural transmission rule did not change during the evolutionary process.

Several of these models included fluctuating selection, which could be due to environmental changes~\citep[reviewed in][]{Aoki2014}. 
We recently studied a model in which each individual can learn a dichotomous phenotype either from a parent, with probability $\rho$, or from a non-parental adult, with probability $1-\rho$ (\autoref{fig:transmission}).
We found that if selection fluctuates between favoring each of the two phenotypes, but on average favored both phenotypes for similar time periods, then a phenotypic polymorphism may be maintained~\citep{Ram2018}.
Furthermore, we found that if the environment changes very rapidly then lower $\rho$ values are likely to evolve, that is, oblique transmission is favored over vertical transmission.

There has been much less theory developed for fluctuating transmission.
Nevertheless, if we assume that social learning can be affected by ecological and demographic factors, such as the frequency of interactions~\citep{VanSchaik2003}, weather~\citep{Phithakkitnukoon2012}, population size or density~\citep{Fischer2015,Aureli1997a}, or stress~\citep{Farine2015}, then it is reasonable that the mode and rate of transmission may fluctuate over time and/or space.
For example, \citet{Webster2008} have demonstrated that minnows (small freshwater shoaling fish), increase their reliance on social learning when predation risk increases. %[MWF-NOTE: non-parental: adults in the parental generation]
%For example, it may be the case that during cold seasons offspring are more likely to stay close to their parents and therefore learn vertically, whereas during warm seasons they are more likely to wander and interact with other adults and therefore learn obliquely.

Here, we analyze a model in which individuals learn from their parents with probability $\rho$ (vertical transmission), and from non-parental adults with probability $1-\rho$ (oblique transmission), where $\rho$ fluctuates over time or space~(\autoref{fig:transmission}).
We find that $\ldots$ (to be continued)  %TODO
We suggest that cultural evolution can be affected by environmental changes in both fitness and transmission fluctuations, and that further empirical and theoretical studies are required to disentangle these effects.

@@

\paragraph{Unconditional migration and symmetric selection.}%[MWF: Why call it unconditional?]

For the simple case of unconditional migration $m_1=m_2=m$ \citep{McPeek1992} after symmetric selection $w_1=1/w_2=w>1$, the recursions~\eqref{eq:migration_model} become
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1'' &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2'' &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

@@

For the stability of the equilibria when $\rho_1=\rho_2=1$, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $a \lambda^2 + b \lambda + c$ with
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w^2+1)}{w}, \quad
c &=  1-2m.
\end{aligned}
\end{equation}
Since $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is greater than one, both equilibria are unstable, and a protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_polymorphism}.

When $\rho_1=1$ and $\rho_2=\rho$, the stability of $(0,0)$ (i.e. fixation in $B$) is determined by the characteristic polynomial with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w+1-\rho+\rho/w)}{w}, \quad
c &=  (1-2m)(w(1-\rho)+\rho).
\end{aligned}
\end{equation}
Again, $a>0$ and $c>0$ and stability can be analyzed by the sign of $a+b+c$.
For $\rho=1$ we already have $a+b+c<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any valid $\rho$, the leading eigenvalue is greater than one, and $(0,0)$ is unstable.

The stability of $(1,1)$ (i.e. fixation in $A$) is determined by the characteristic polynomial with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{1-m}{w}(1 + w(1-\rho) +w^2 \rho), \quad
c &= \frac{1-2m}{w} (w\rho+1-\rho),
\end{aligned}
\end{equation}
and real roots. 
Again, $a>0$, $b<0$, and $c>0$, and if $\rho=1$ then $a+b+c<0$, as before.
However, for $\rho=0$ we have $a+b+c = \frac{m}{w}(w-1)>0$, in which case the stability is determined by $2a+b=\frac{m(w+1)+w-1}{w}$, the derivative of the polynomial at $\lambda=1$, which is positive, and therefore both eigenvalues occur in $(0,1)$ and fixation is stable. %[MWF: Are the roots real?]

@@

The values of $a+b+c$ at $m=0$ and $m=1$ are $-\frac{\rho(w-1)^2}{w}<0$ and $\frac{(1-\rho)(w-1)}{w}>0$, respectively, and $a+b+c$ is linear in $m$, so that we have $a+b+c>0$ when 
\begin{equation}
m > \hat m = \frac{\rho (w-1)}{\rho (w-1) +1-\rho}.
\end{equation}
In addition, $2a+b$ is linear in $m$ and equal to 2 when $m=1$; it is equal to zero when $m=(w-1)(w\rho-1)/(w(w-1)\rho+w+1) < \hat m$, and therefore \eqref{eq:migration_model_unconditional_symmetric_condition_m} 
is sufficient for stability of fixation of $A$.
%[MWF: (22, also 27)]

>>>>

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

@@

Important modes of cultural transmission are vertical (from parent to offspring), horizontal (between peers), and oblique (from adults in the parental generation to offspring) transmission. 
\citet[ch.~3]{Cavalli-Sforza1981} first introduced models in which a specific trait is transmitted either vertically or obliquely.
Mathematical models of cultural evolution may include one or more modes of cultural transmission and in most of these models the mode of transmission by individuals is fixed; that is, it does not vary over time.
For example, \cite{Feldman1996}, \cite{Wakano2004}, \cite{Aoki2005}, and \citet{McElreath2008} focused on competition between individual learning, innate learning, and social learning; \citet{Fogarty2017} compared the effects of different oblique mechanisms (random, success-biased, best-of-k, one-to-many) on the cultural richness and diversity of the population; and \citet{Aoki2012} modeled scenarios in which individual and social learning occur during separate stages in life.
In these studies, the cultural transmission rule did not change during the evolutionary process.

Several of these models included fluctuating selection, which could be due to environmental changes~\citep[reviewed in][]{Aoki2014}. 
We recently studied a model in which each individual can learn a dichotomous phenotype either from a parent, with probability $\rho$, or from an adult in the parental generation, with probability $1-\rho$ (\autoref{fig:transmission}).
We found that if selection fluctuates between favoring each of the two phenotypes, but on average favored both phenotypes for similar time periods, then a phenotypic polymorphism may be maintained~\citep{Ram2018}.
Furthermore, we found that if the environment changes very rapidly then lower $\rho$ values are likely to evolve, that is, oblique transmission is favored over vertical transmission.

There has been much less theory developed for fluctuating transmission.
Nevertheless, if we assume that social learning can be affected by ecological and demographic factors, such as the frequency of interactions~\citep{VanSchaik2003}, weather~\citep{Phithakkitnukoon2012}, population size or density~\citep{Fischer2015,Aureli1997a}, or stress~\citep{Farine2015}, then it is reasonable that the mode and rate of transmission may fluctuate over time and/or space.
For example, \citet{Webster2008} have demonstrated that minnows (small freshwater shoaling fish), increase their reliance on social learning when predation risk increases. 
%For example, it may be the case that during cold seasons offspring are more likely to stay close to their parents and therefore learn vertically, whereas during warm seasons they are more likely to wander and interact with other adults and therefore learn obliquely.

Here, we analyze a model in which individuals learn from their parents with probability $\rho$ (vertical transmission), and from adults in the parental generation with probability $1-\rho$ (oblique transmission), where $\rho$ fluctuates over time or space~(\autoref{fig:transmission}).
We find that $\ldots$ (to be continued)  %TODO
We suggest that cultural evolution can be affected by environmental changes in both fitness and transmission fluctuations, and that further empirical and theoretical studies are required to disentangle these effects.

@@

\paragraph{Unconditional migration and symmetric selection.}

For the simple case of unconditional migration $m_1=m_2=m<1/2$ \citep{McPeek1992} after symmetric selection $w_1=1/w_2=w>1$, the recursions~\eqref{eq:migration_model} become
\begin{equation}\begin{aligned} \label{eq:migration_model_unconditional_symmetric}
x_1'' &= (1-m)x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + m x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big), \\
x_2'' &= m x_1\Big(\rho_1 \frac{w}{\overline w_1} x_1 + 1-\rho_1 \Big) + (1-m) x_2\Big(\rho_2 \frac{1/w}{\overline w_w} x_2 + 1-\rho_2 \Big).
\end{aligned}
\end{equation}

@@

For the stability of the equilibria when $\rho_1=\rho_2=1$, the characteristic polynomial of the Jacobian is the same for both fixation equilibria $(0,0)$ and $(1,1)$ and is given by $f(\lambda) = a \lambda^2 + b \lambda + c$ with
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -\frac{(1-m)(w^2+1)}{w}, \quad
c &=  1-2m.
\end{aligned}
\end{equation}
First, the discriminant $\Delta = b^2-4ac$ is positive for any real $m$, because, when writing it as a quadratic polynomial in $m$, $\Delta(m)=\frac{1}{w^{2}} \big(m^{2} (w^2+1)^2 - 2 m (w^2-1)^2 + (w^2-1)^2\big)$, this polynomial is positive at $m=0$ and $m=1/2$ and has a negative discriminant $-16w^2(w^2-1)^2$.
Therefore, the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two positive roots. \\
Second, as $a>0$, $b<0$, and $c>0$, and because $a+b+c=-\frac{1-m}{w}(w-1)^2<0$, the leading eigenvalue is greater than one, both equilibria are unstable, and a protected polymorphism exists and is given by~\eqref{eq:migration_model_unconditional_symmetric_polymorphism}.

When $\rho_1=1$ and $\rho_2=\rho$, the stability of $(0,0)$ (i.e. fixation in $B$) is determined by the characteristic polynomial $f(\lambda)$ with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1 - m) (w + 1-\rho + \rho/w), \quad
c &=  (1-2m)(w(1-\rho)+\rho).
\end{aligned}
\end{equation}

First, the discriminant is 
$$
\Delta = (1 - m)^2 (w + 1-\rho + \rho/w)^2 - 4(1-2m)(w(1-\rho)+\rho),
$$
which is positive for $m=0$ and $m=1/2$:
\begin{equation}\begin{aligned}
\Delta(m=0) = \Big(\frac{(w + \rho) (w-1)}{w}\Big)^2 \\
\Delta(m=1/2) = \Big(\frac{w^2 + w(1-\rho) + \rho}{2 w}\Big)^2.
\end{aligned}\end{equation}
In addition, if we write $\Delta$ as a polynomial in $m$, its discriminant is negative -- $- 16 (w + \rho)^2 (w - 1)^2 \big(w (1 - \rho) + \rho \big) / w^2$, and therefore the discriminant $\Delta$ is positive for any $0 \le m \le 1/2$ and $f(\lambda)$ has two real roots. \\
Second, $a>0$, $b<0$, and $c>0$ and stability can be analyzed by the sign of $a+b+c$.
For $\rho=1$ we already have $a+b+c<0$.
For $\rho=0$ we have $a+b+c=-m(w-1)<0$.
Finally, $a+b+c$ is a linear function of $\rho$ and therefore $a+b+c<0$ for any valid $\rho$, the leading eigenvalue is greater than one, and $(0,0)$ is unstable.

The stability of $(1,1)$ (i.e. fixation in $A$) is determined by the characteristic polynomial with coefficients
\begin{equation}
\begin{aligned}
a &= 1, \quad
b &= -(1-m)(w \rho + 1 -\rho + 1/w), \quad
c &= (1-2m) \big((1-\rho)/w + \rho\big),
\end{aligned}
\end{equation}
%[MWF: Are the roots real?]
Again, $a>0$, $b<0$, and $c>0$, and if $\rho=1$ then $a+b+c<0$, as before.
However, for $\rho=0$ we have $a+b+c = \frac{m}{w}(w-1)>0$, in which case the stability is determined by $2a+b=\frac{m(w+1)+w-1}{w}$, the derivative of the polynomial at $\lambda=1$, which is positive, and therefore both eigenvalues occur in $(0,1)$ and fixation is stable. %

@@

The values of $a+b+c$ at $m=0$ and $m=1$ are $-\frac{\rho(w-1)^2}{w}<0$ and $\frac{(1-\rho)(w-1)}{w}>0$, respectively, and $a+b+c$ is linear in $m$, so that we have $a+b+c>0$ when 
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m2}
m > \hat m = \frac{\rho (w-1)}{\rho (w-1) +1-\rho}.
\end{equation}
In addition, $2a+b$ is linear in $m$ and equal to 2 when $m=1$; it is equal to zero when $m=(w-1)(w\rho-1)/(w(w-1)\rho+w+1) < \hat m$, and therefore (\ref{eq:migration_model_unconditional_symmetric_condition_m}, also \ref{eq:migration_model_unconditional_symmetric_condition_m2})
is sufficient for stability of fixation of $A$.
